---
title: "GOexpress version **2.x.x** - User Guide"
author:
    -   name: Kévin Rue-Albrecht
        email: kevinrue67@gmail.com
        affiliation:
            - School of Veterinary Medicine, University College Dublin, Ireland
            - Nuffield Department of Medicine, University of Oxford, UK
date: "`r doc_date()`"
package: "`r pkg_ver('GOexpress')`"
abstract:
    This package contains methods to visualise gene expression data, typically
    microarray or RNA-sequencing, and offers a supervised approach to identify
    Gene Ontology (GO) categories of genes that best classify two or more
    predefined groups of samples.
    This vignette presents version **2.x.x** of the `r Biocpkg("GOexpress")`
    package, that introduces classes and methods (_S3_ and _S4_) for better
    interoperability with other _Bioconductor_ packages.
vignette: >
    %\VignetteIndexEntry{GOexpress version **2.x.x** - User Guide}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output:
    BiocStyle::pdf_document
bibliography:
    GOexpress.bib
nocite: | 
  @GOexpress
---

```{r setup, include=FALSE}
stopifnot(
    requireNamespace("pander")
)
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
optChunkDefault <- opts_chunk$get()
```

# Toward version **2.x.x**

This vignette introduces an updated implementation of the original features
that introduces classes and methods (*S3* and *S4*) for better
interoperability and integration with other _Bioconductor_ packages.

The motivation for this transition was initially prompted by user feedback
(thanks!) pointing out how existing functions
were restricted to objects of the `ExpressionSet` class,
in contrast to several classes derived from the more recent
`SummarizedExperiment` class.

# Demonstration environment

To set up the environment for this vignette, let us first load
the `r Biocpkg("GOexpress")` package:

```{r require_GOexpress, message=FALSE}
stopifnot(require(GOexpress))
```

We can then import the demonstration data set:

```{r data_AlvMac}
data("AlvMac")
```

This data set is an `ExpressionSet` subset of
`r as.numeric(nrow(AlvMac))` features (_i.e._, genes) and
`r as.numeric(ncol(AlvMac))` samples,
from a larger data set:

```{r AlvMac}
AlvMac
```

The data set includes phenotype metadata for each sample:

```{r varLabels}
varLabels(AlvMac)
```

**Note:**
The `Time` field is a _factor_,
while the `Timepoint` field stores the _numeric_ value (in hours).

# Random forest

A random forest (RF) is a powerful supervised  machine learning (ML) approach
that builds many classification trees (hence, 'forest')
using random subsamples of predictors (here, genes or other features
such as microarray probe sets) to score those predictors according to their
capacity to classify samples from different predefined groups
(*e.g.*, experimental treatment).

The _S3_ method dispatch introduced for versions **2.x.x** of
`r Biocpkg("GOexpress")` now enable seamless support of a wider range
of common _Bioconductor_ classes (and classes that inherits from them):

* `r Biocpkg("DESeq2")` `DESeqDataSet` and `DESeqTransform` classes,
* the `r Biocpkg("edgeR")` `DGEList` class,
* the `r Biocpkg("SummarizedExperiment")` class of the eponym package,
    widely extended by other _Bioconductor_ packages
    (*e.g.*, `SingleCellExperiment`)
* the `r Biocpkg("Biobase")` `ExpressionSet` class originally supported.

Notably, the only additional argument absolutely required
is the name of a phenotype factor in `varLabels(AlvMac)`,
that will be used as a response vector to run the _randomForest_ algorithm
in _supervised mode_:

```{r randomForest}
rf <- randomForest(AlvMac, pheno="Treatment", do.trace = 100, verbose=TRUE)
```

**Note:**
The `GOexpress::randomForest` implementation can display informative messages
to indicate the count of samples within each phenotype factor level;
while the underlying `randomForest::randomForest` implementation
can display 'out-of-bag' error rates every `do.trace` trees.
The former is disabled by default with `verbose=FALSE`,
while the latter is enabled by default (`do.trace=100`)
and can be disabled with `do.trace=FALSE`.

Notably,
the `GOexpress::randomForest` call returns _exactly_ the output of the
underlying `randomForest::randomForest` method. This means that the same
downstream methods may be applied; for instance the `plot` method:

```{r plot_rf, results='hold'}
op <- par(no.readonly = TRUE)
layout(matrix(c(1,2),nrow=1), width=c(4,1)) 
par(mar=c(5,4,4,0)) # No margin on the right side
plot(rf, ylim = c(0,1), main = "OOB error by iteration - AlvMac")
par(mar=c(5,0,4,2)) # No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("top", colnames(rf$err.rate),col=1:4,cex=0.8,fill=1:4)
par(op)
```

**Reference**:
This StackOverflow [thread](http://stackoverflow.com/q/20328452)
entitled _Legend for Random Forest Plot in R_.

From a biological interpretation perspective, in the above figure,
the higher out-of-bag (OOB) error rate for classification of
_M. tuberculosis_ and _M. bovis_ samples can be largely attribued
to the similar host macrophage response to the closely
related bacterial species.
Specifically, those error rates and the associated confusion matrix (below)
indicate that the supervised machine learning classification
often mistakes one infection for the other;
in contrast, control (*i.e.*, uninfected) samples are much more
accurately classified:

```{r confusion, results='asis', echo=FALSE}
pander::pandoc.table(rf$confusion)
```

# Citation

```{r citation}
citation(package="GOexpress")
```

# Session info

This vignette was built in the following environment:

```{r sessionInfo}
sessionInfo()
```

# References {#References}
